---
title: "Mt1 2018"
author: "Irlanda Ayon-Moreno"
date: "5/4/2018"
output: pdf_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r message=FALSE}
library(dplyr)
library(ggplot2)
```

Explained Errors
================

```{r}
dat <- data.frame(
  first = c("Jon", "Arya", "Tyrion", "Daenerys", "Yara"),
  last = c("Snow", "Stark", "Lannister", "Targaryen", "Greyjoy"),
  gender = c("male", "female", "male", "female", "female"),
  title = c("lord", "princess", "master", "khaleesi", "princess"),
  gpa = c(2.8, 3.5, 2.9, 3.7, NA),
  stringsAsFactors = FALSE
)
dat
```

## value of 'first' for maximum 'gpa'

```{r}
max_gpa <- max(dat$gpa, na.rm = TRUE)
max_gpa
```

```{r}
which_max_gpa <- dat$gpa = max_gpa
```

```{r}
# corrected
which_max_gpa <- dat$gpa == max_gpa
which_max_gpa
```

```{r}
dat$first(which_max_gpa)
```

```{r}
# corrected
dat$first[which_max_gpa]
```

## gpa of title lord

```{r}
dat[ , c("title", "gpa")]
```

```{r}
dat[ , "title"]
```

```{r}
dat[ , c("title", "gpa")] == "lord"
```

```{r}
dat[ , "title"] = "lord"
dat
```

```{r}
dat$gpa[dat[ , "title"] = "lord"]
```

```{r}
# corrected
dat$gpa[dat[ , "title"] == "lord"]
```

## median gpa of each gender

```{r}
which_males <- dat$gender == 'male'
which_females <- dat$gender == 'female'
```

```{r}
# with mistake
median_females <- median(dat$gpa[which_males])
median_males <- median(dat$gpa[which_males])
c(median_females, median_males)
```

```{r}
# corrected 
median_females <- median(dat$gpa[which_females], na.rm = TRUE)
median_males <- median(dat$gpa[which_males])
c(median_females, median_males)
```

Command Outputs
================

```{r}
student <- list(
  name = "Anakin Skywalker",
  gpa = 4,
  major_minor = c(major1 = "jedi studies", major2 = "sith studies", minor = "galactic policies"),
  grades = data.frame(
    course = c("force-101", "podracing", "light-sabers"),
    score = c(9.3, 10.0, 8.5),
    stringsAsFactors = FALSE
  )
)
student
```

```{r}
length(student$major_minor)
```

```{r}
student$gpa < 2.5
```

```{r}
names(student$major_minor)
```

```{r}
student$grades[2]
```

```{r}
student$grades[[2]]
```

```{r}
student$grades[[2]][2]
```

```{r}
rep(student$grades[[2]][2], student$gpa)
```

```{r}
student$grades %>% arrange(score)
```

Data Types
===========

```{r}
uno <- c(TRUE, FALSE)
uno
typeof(uno)
```

```{r}
dos <- uno + 1L
dos
typeof(dos)
```

```{r}
tres <- c(uno, dos, "3.0", "4.0")
tres
typeof(tres)
```

```{r}
cuatro <- as.factor(tres)
cuatro
typeof(cuatro)
```

```{r}
uno[1]
dos[uno[1]]
cinco <- tres[dos[uno[1]]]
cinco
typeof(cinco)
```

Markdown
=========

## Examples

Text in italics:

*italics*

Text in bold:

**bold**

Text that is associated to a (hyper)link:

[link title](www.url.com)

Text in code format:

- code chunck: Make a code chunk with three back ticks followed
by an r in braces. End the chunk with three back
ticks
```{r}
four <- 4
```

- inline code: use back ticks around the code. include an r after the first back tick so that the code updates itself.  


      `four` or `r four`  
    
How to write unordered bulleted list:

Use - or + or * & end each line with 2 spaces

How to write ordered bulleted list:

Use numbers & end each line with 2 spaces

Vector Subsetting
=================

```{r}
lord <- c("v", "o", "l", "d", "e", "m", "o", "r", "t")
```

```{r}
names(lord) = 1:9
lord
```

```{r}
lord[length(lord)]
```

```{r}
as.logical(-1:-4)
lord[as.logical(-1:-4)]
```

```{r}
lord[seq(from = 1, to = 9, by = 2)]
```

```{r}
lord[lord != "e" | lord == "o"]
```

```{r}
lord[3:1]
lord[3:1] == lord[3:1]
```

## Write code that returns the elements: "d" "o" "o" "r"

```{r}
lord[c(4, 2, 7, 8)]
```

Explain Commands
=================

## What does `sum(!is.finite(x))` say about `x`?

```{r}
x <- seq(from = 0, to = 12, by = 2)
!is.finite(x)
sum(!is.finite(x))

x <- c(0, 1, 2, 3, 4, pi/0, 1/0) # a non-zero number divided by 0 creates infinity
!is.finite(x)
sum(!is.finite(x))

# Tells the number of infinite values.
# 0 if x is finite
```

## COnsider a numeric vector `x`. What type of subsetting operaations are involved in the command: `x[-which(x > 0)]`?

```{r}
# numeric subsetting

x <- c(-4, -3, -2, -1, 0, 1, 2, 3, 4)
x > 0
which(x > 0)
-which(x > 0)
x[-which(x > 0)]
```

## A data frame `A` and a matrix `B` contain exactly the same tabular data. The first column of `A` and `B` is named `Col1`. Why does `B$Col1` return an error?

```{r}
B <- matrix(c(1,2,3, 11,12,13), nrow = 2, ncol = 3, byrow = TRUE,
            dimnames = list(c("row1", "row2"),
                            c("Col1", "Col2", "Col3"))
)
B

A <- data.frame(B)
A
```

```{r}
# $ is used on data frames (lists)
# [] may be uces on data frames (lists) & On Matrices
#     [row idex, column index]

A$Col1
B$Col1
A[ , 1]
B[ , 1]
```

dplyr
======

```{r}
autos <- data.frame(
  Model = as.character(c("pontiac firebird", "pontiac safari", NA, "opel 1900", "peugeot 304", "volkswagen model 111", NA, "datsun p1510")),
  MPG = as.double(c(19.0, NA, 30.0, NA, 30.0, 27.0, 35.0, NA)),
  Cylinders = as.integer(c(6, 8, rep(4, 6))),
  Weight = as.integer(c(NA, 5140, 2065, 2123, 2074, 1834, 1613, 2130)),
  Accelerate = as.double(c(15.0, 12.0, 14.5, NA, 19.5, NA, 18.0, 14.5)),
  Origin = as.character(c(rep("American", 2), rep("European", 4), rep("Japanese", 2)))
)
autos
```

## Display the `Model` and `Weight` of all `European` cars with `MPG` larger than 25. The output should be ordered from largest to smallest (in weight).

```{r}
autos %>%
  filter(Origin == "European" & MPG > 25) %>%
  select(c(Model, Weight)) %>%
  arrange(desc(Weight))
```

ggplot2
========

## Create a scatterplot of `mpg` and `hp`, with the size of the dots depending on `cyl`, and faceting by `transmission`.

```{r}
dataset <- data.frame(
  row.names = c("Mazda RX4", "Mazda RX4 Wag", "Dotsun 710", "Hornest 4 Drive", "Hornest Sportabout"),
  mpg = c(21, 21, 22.8, 21.4, 18.7),
  cyl = c(6, 6, 4, 6, 8),
  hp = c(110, 110, 93, 110, 175),
  transmission = c(rep("automatic", 3), rep("manual", 2))
)
dataset
```

```{r}
ggplot(data = dataset, mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(size = cyl)) +
  facet_grid(.~transmission)
```

