---
title: "Mt1 2017"
author: "Irlanda Ayon-Moreno"
date: "5/4/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r message=FALSE}
library(dplyr)
library(ggplot2)
```


Command Outputs
===============

```{r}
x <- 1:5
y <- letters[1:5]
names(y) <- 1:5
```

## What is the output of `y[x <= 2]`?

```{r}
y[x <= 2]
```

## What is the output of `y[x[5]]`?

```{r}
y[x[5]]
```

## What is the output of `y[!(x < 3)]`?

```{r}
y[!(x < 3)]
```

## What is the output of `y[x/x]`?

```{r}
y[x/x]
```

## What is the output of `y[x[-2][3]]`?

```{r}
y[x[-2][3]]
```

## What is the output of `x[(y != "c") & (y != "a")]`?

```{r}
x[(y != "c") & (y != "a")]
```

## What is the output of `x[(y == "c") | (y == "a")]`?

```{r}
x[(y == "c") | (y == "a")]
```

## Do these commands return the numeric vector given by: 1 2 3 4 5?

* `c(1 2 3 4 5)`  
* `1:5`  
* `list(1, 2, 3, 4, 5)`  
* `seq(from = 1, to = 5, by = 1)`  
* `[1, 2, 3, 4, 5]`  
* `c("1", "2", "3", "4", "5")`
* `1 + c(0, 1, 2, 3, 4)`  
* `sqrt(1, 4, 9, 16, 25)`

```{r}
c(1 2 3 4 5) # no
1:5 # yes
list(1, 2, 3, 4, 5) # no
seq(from = 1, to = 5, by = 1) # yes
[1, 2, 3, 4, 5] # no
c("1", "2", "3", "4", "5") # no
1 + c(0, 1, 2, 3, 4) # yes
sqrt(1, 4, 9, 16, 25) # no
```

Data Types
===========

```{r}
x <- c(1.1, 2.2, 3.3, 4.4)
x
typeof(x)
```

```{r}
y <- (x == 1.1)
y
typeof(y)
```

```{r}
z <- y + 0
z
typeof(z)
```

```{r}
w <- c(x, "5.5")
w
typeof(w)
```

```{r}
xyz1 <- c(x, y, z, w[1])
xyz1
typeof(xyz1)
```

Files
======

## True or False?

**"\t"**, **";"**, **" "**, and **","** are charcters commonly used as field-delimiters

A file format is a way of interpreting the bytes in a file

The CSV format name stands for Common Standard Value

A CSV format is a special case of a spreadsheet format

In a CSV file, there must be a single header line containing the names of the fields

Binary formats tend to provide smaller files and faster access speeds

One disadvantage of plain text files is their lack of standard structure

Data stored in a binary format can be accessed using a text editor


ggplot2
========

```{r, echo = FALSE}
sw <- data.frame(
  name = c("Anakin", "Padme", "Luke", "Leia"),
  gender = c(rep(c("male", "female"), 2)),
  height = c(1.88, 1.65, 1.72, 1.5),
  weight = c(84, 45, 77, 49)
)
```

```{r}
sw
```

## Create a scatter plot of `height` and `weight`. Distinguish the points using a person's `gender` and labeling each point with their `name`. Finally, add a title.

```{r}
ggplot(data = sw, mapping = aes(x = height, y = weight)) +
  geom_point(aes(shape = gender)) +
  geom_text(aes(label = name), vjust = "inward") +
  ggtitle("Scatter Diagram of Height and Weight")
```

## Create a scatterplot of `mpg` and `hp`, with the size of the dots depending on `cyl`, the shape of the dots depending on `transmission`, and the color of the dots depending on `transmission`. Facet by `transmissionm` and include a title.

```{r, echo = FALSE}
coches <- data.frame(
  row.names = c("Mazda RX4", "Mazda RX4 Wag", "Dotsun 710", "Hornest 4 Drive", "Hornest Sportabout"),
  mpg = c(21, 21, 22.8, 21.4, 18.7),
  cyl = c(6, 6, 4, 6, 8),
  hp = c(110, 110, 93, 110, 175),
  transmission = c(rep("automatic", 3), rep("manual", 2))
)
```

```{r}
coches
```

```{r}
ggplot(data = coches, mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(size = cyl, shape = transmission, color = transmission)) +
  facet_grid(.~transmission) +
  ggtitle("MPG and HP by Transmission")
```

dplyr
======

```{r}
sw
```

## Which of the following commands returns the number of females and males?

* `sw %>% select(gender) %>% count()`  
* `sw %>% count(gender)`  
* `sw %>% group_by(gender) %>% count()`  
* `sw %>% group_by(gender) %>% summarise()`


```{r}
sw %>% select(gender) %>% count() # no
sw %>% count(gender) # yes
sw %>% group_by(gender) %>% count() # yes
sw %>% group_by(gender) %>% summarise() # no
```

## Which of the following commands returns the data (rows) of male individuals?

* `sw %>% select(gender == "male")`  
* `sw %>% filter(gender == "male")`  
* `sw %>% group_by(gender == "male")`  
* `sw %>% filter(by == "male")`


```{r}
sw %>% select(gender == "male") # no
sw %>% filter(gender == "male") # yes
sw %>% group_by(gender == "male") # no
sw %>% filter(by == "male") # no
```

