---
title: "W5"
author: "Irlanda Ayon-Moreno"
date: "4/10/2018"
output: pdf_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

### dplyr

**filter**: keep rows matching criteria

**select**: pick columns by name

**arrange**: reorder rows

**mutate**: add new variables

**summarise**: reduce variables to values


#### Structure  
* 1st argument is a data frame  
* next arguments say what to do with the data frame  
* always returns a data frame

### ggplot2

Automated things in ggplot2:
* axis labels  
* legends  
* color for points  
* grid lines  
* axis tick marks  

```{r}
name <- c("Luke Skywalker", "Leia Skywalker", "Obi-Wan Kenobi", "Han Solo", "R2-D2", "C-3PO", "Yoda", "Chewbacca")
gender <- c("male", "female", rep("male", 6))
height <- c(1.72, 1.5, 1.82, 1.8, 0.96, 1.67, 0.66, 2.28)
weight <- c(77, 49, 77, 80, 32, 75, 17, 112)
jedi <- c("jedi", "no_jedi", "jedi", rep("no_jedi", 3), "jedi", "no_jedi")
species <- c(rep("human", 4), rep("droid", 2), "yoda", "wookie")
weapon <- c(rep(c("lightsaber", "blaster"), 2), rep("unarmed", 2), "lightsaber", "bowcaster")
starwars_df <- data.frame(name, gender, height, weight, jedi, species, weapon)
starwars_df
```

```{r}
ggplot(data = starwars_df) +
  geom_point(mapping = aes(x = height, y = weight, color = jedi))
```

**A graphic is the mapping from data to aesthetic attributes (color, shape, size) of geometric objects (points, lines, bars, etc)**
